---
title: "Bayesian multilevel modelling and estimation of uncertainty in speaker variation"
author: "Stefano Coretta"
format: revealjs
---

```{r}
#| label: setup

library(tidyverse)
theme_set(theme_light())

library(coretta2019eng)
data("eng_durations")
library(brms)
library(tidybayes)

library(MASS)

```

## {background="var(--inverse)"}

::: r-fit-text
What does it mean to model speaker variation?
:::

. . .

::: r-fit-text
It means to model the population of speakers.
:::

## English voicing effect

```{r}
#| label: ve

ve_plot <- eng_durations |> 
  ggplot(aes(vowel_ipa, v1_duration, colour = voicing)) +
  scale_color_brewer(type = "qual") +
  labs(
    x = "Vowel",
    y = "Vowel duration (ms)",
    colour = "Consonant\nvoicing",
    caption = "15 speakers from Manchester"
  )
ve_plot
```

## English voicing effect

```{r}
#| label: ve-2

ve_plot +
  geom_jitter(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.2, seed = 4127))
```

## Four speakers

```{r}
#| label: ve-3

eng_durations |> 
  filter(speaker %in% c("en01", "en02", "en10", "en12")) |> 
  ggplot(aes(vowel_ipa, v1_duration, colour = voicing)) +
  geom_jitter(alpha = 0.75, position = position_jitterdodge(jitter.width = 0.2, seed = 4127)) +
  facet_wrap(vars(speaker)) +
  scale_color_brewer(type = "qual") +
  labs(
    x = "Vowel",
    y = "Vowel duration (ms)",
    colour = "Consonant\nvoicing"
  )
```

## Modelling the distribution of speakers

(Bayesian) multilevel regression models

. . .

aka hierarchical

. . .

aka mixed-effects

## Bayesian inference

$$
P(h \mid d) \propto P(h) \cdot P(d \mid h)
$$

where:

- $P(h)$ is the *prior probability* of $h$.
- $P(d \mid h)$ is the probability of $d$ given $h$ (aka *likelihood*).
- $P(h \mid d)$ is the *posterior probability* of $h$ given $d$.

## Bayesian updating

```{=html}
<iframe width="100%" height="500" src="https://nanx.shinyapps.io/conjugate-normal-umkv/"></iframe>
```

::: aside
App by [Nan Xiao](https://nanx.me), code on <https://github.com/nanxstats/conjugate-normal-umkv>
:::


## Regression model

$$
\begin{aligned}
v_i & \sim LogNormal(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot voi_i
\end{aligned}
$$

## Multilevel regression model: varying terms

$$
\begin{aligned}
v_i & \sim LogNormal(\mu_i, \sigma)\\
\mu_i & = \beta_0 + u_{0j} + (\beta_1 + u_{1j}) \cdot voi_{ij}\\
\begin{pmatrix}u_{0j}\\u_{1j}\end{pmatrix}& \sim \mathcal{MVN} \left(\begin{pmatrix}0\\0\end{pmatrix},\begin{pmatrix}\tau_0^2 & \rho\tau_0\tau_1\\\rho\tau_0\tau_1 & \tau_1^2\end{pmatrix}\right)
\end{aligned}
$$

. . .

- $j$: speaker index

- $u_{0j}$: speaker-specific intercept adjustments

- $u_{1j}$: speaker-specific slope adjustments (i.e. effect of voicing)

## Varying terms matrix

$$
\begin{aligned}
\begin{pmatrix}u_{0j}\\u_{1j}\end{pmatrix}& \sim \mathcal{MVN} \left(\begin{pmatrix}0\\0\end{pmatrix},\begin{pmatrix}\tau_0^2 & \rho\tau_0\tau_1\\\rho\tau_0\tau_1 & \tau_1^2\end{pmatrix}\right)
\end{aligned}
$$

- $\mathcal{MVN}$: multivariate normal (Gaussian)

- $\tau_0^2$: variance among speaker-specific intercepts

- $\tau_1^2$: variance among speaker-specific slopes

- $\rho$: correlation between speaker-specific intercepts ($u_{0j}$) and slopes ($u_{1j}$)

## Variance-covariance matrix

$$
\begin{aligned}
\begin{pmatrix}\tau_0^2 & \rho\tau_0\tau_1\\\rho\tau_0\tau_1 & \tau_1^2\end{pmatrix} = \begin{pmatrix}\tau_0 & 0\\0 & \tau_1\end{pmatrix}
\begin{pmatrix}1 & \rho\\\rho & 1\end{pmatrix}
\begin{pmatrix}\tau_0 & 0\\0 & \tau_1\end{pmatrix}
\end{aligned}
$$

- $\tau_0$: standard deviation of speaker-specific intercepts

- $\tau_1$: standard deviation of speaker-specific slopes

- $\rho$: correlation between speaker-specific intercepts ($u_{0j}$) and slopes ($u_{1j}$)

## $\tau_0$ example

```{r}
#| label: tau0

x <- seq(-0.6, 0.6, by = 0.01)
y <- dnorm(x, sd = 0.2)

tau0 <- ggplot() +
  aes(x, y) +
  geom_vline(xintercept = 0, colour = "gray") +
  geom_line() +
  labs(
    title = expression(tau[0] == 0.2)
  )

tau0
```

## $\tau_0$ example: 15 speakers (sample A)

```{r}
#| label: tau0-15

set.seed(42368)
u <- rnorm(15, 0, 0.2)

tau0 +
  geom_point(aes(u, 0), colour = "indianred", size = 3, shape = 1) +
  labs(
    x = expression(u[0] ~ "(intercept adjusment)"),
    y = "density",
    caption = "Each circle is a speaker."
  )
```

## $\tau_0$ example: 15 speakers (sample B)

```{r}
#| label: tau0-15

set.seed(423610)
u <- rnorm(15, 0, 0.2)

tau0 +
  geom_point(aes(u, 0), colour = "indianred", size = 3, shape = 1) +
  labs(
    x = expression(u[0] ~ "(intercept adjusment)"),
    y = "density",
    caption = "Each circle is a speaker."
  )
```

## $\rho$ example

```{r}
#| label: rho
#| warning: false

set.seed(123)

# Parameters
n_speakers <- 15
tau_0 <- 0.2
tau_1 <- 0.04

# Nine correlations
rho_values <- c(-0.9, -0.6, -0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.9)

# Simulate speaker-specific intercept and slope adjustments
sim_data <- lapply(rho_values, function(rho) {
  
  Sigma <- matrix(
    c(tau_0^2, rho * tau_0 * tau_1,
      rho * tau_0 * tau_1, tau_1^2),
    nrow = 2
  )
  
  u <- mvrnorm(
    n = n_speakers,
    mu = c(0, 0),
    Sigma = Sigma
  )
  
  data.frame(
    speaker = 1:n_speakers,
    u_0 = u[, 1],
    u_1 = u[, 2],
    rho = rho
  )
}) |> 
  bind_rows()

rho_plot <- ggplot(sim_data, aes(x = u_0, y = u_1)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "gray") +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "gray") +
  facet_wrap(
    ~ rho,
    nrow = 3,
    ncol = 3,
    labeller = labeller(
      rho = function(x) paste0("ρ = ", x)
    )
  ) +
  labs(
    x = expression(u[0] ~ "(intercept adjustment)"),
    y = expression(u[1] ~ "(slope adjustment)"),
    caption = "Each circle is a speaker."
  )

rho_plot +
  geom_point(size = 2.5, colour = NA, shape = 1)
```

## $\rho$ example

```{r}
#| label: rho-2

rho_plot +
  geom_point(size = 2.5, colour = "indianred", shape = 1)
```


## Fitting the model

```{r}
#| label: ve-bm

ve_bm <- brm(
  v1_duration ~ voicing + (voicing | speaker),
  family = lognormal,
  data = eng_durations,
  cores = 4,
  seed = 2136,
  file = "data/cache/ve_bm"
)

```

## Multilevel hyper-parameters summary

```{r}
#| label: ve-bm-summary

summary(ve_bm, prob = 0.9)

```

## Intercept of vowel duration

```{r}
#| label: int

ve_draws <- as_draws_df(ve_bm) |> 
  mutate(
    int_exp = exp(b_Intercept),
    voi_r = exp(b_voicingvoiced),
    voiceless = int_exp,
    voiced = exp(b_Intercept + b_voicingvoiced)
  )

ve_draws |> 
  ggplot(aes(int_exp)) +
  geom_density(fill = "aquamarine4", colour = NA) +
  labs(
    x = "Intercept of V duration (ms)"
  )
```

## Effect of voicing

```{r}
#| label: voi-ratio

ve_draws |> 
  ggplot(aes(voi_r)) +
  geom_vline(xintercept = 1, colour = "deepskyblue4") +
  geom_density(fill = "tomato3", colour = NA) +
  labs(
    x = "Voicing effect ratio"
  )
```
## Expected values of vowel duration

```{r}
#| label: ve-post
#| warning: false

ve_draws |> 
  dplyr::select(voiceless, voiced) |> 
  pivot_longer(c(voiceless, voiced), names_to = "voicing") |> 
  mutate(voicing = factor(voicing, c("voiceless", "voiced"))) |> 
  ggplot(aes(value, fill = voicing)) +
  geom_density(alpha = 0.85, colour = NA) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Expected V duration (ms)"
  )
```

## By-speaker adjustments to intercept

```{r}
#| label: r-int

ve_speaker <- ve_bm |> 
  spread_draws(r_speaker[speaker,coef], b_voicingvoiced) |> 
  mutate(
    bs_voicing = b_voicingvoiced + r_speaker
  )

ve_speaker |> 
  filter(coef == "Intercept") |> 
  ggplot(aes(r_speaker, reorder(speaker, r_speaker, median))) +
  geom_vline(xintercept = 0) +
  stat_halfeye() +
  labs(
    x = "By-speaker intercept adjustment (log)",
    y = "Speaker"
  )
```

## $\tau_0$: SD of speaker-specific adjustments

```{r}
#| label: tau0-post

ve_draws |> 
  ggplot(aes(sd_speaker__Intercept)) +
  geom_density(fill = "forestgreen", colour = NA)
```

## $u_0$ posterior probability (20 draws)

```{r}
#| label: tau0

set.seed(5364)

expand_grid(
  sd = sample(ve_draws$sd_speaker__Intercept, size = 20),
  u = seq(-1, 1, by = 0.01)
) |> 
  mutate(
    dens = dnorm(u, 0, sd = sd)
  ) |> 
  ggplot(aes(u, dens, group = sd)) +
  geom_path(alpha = 0.25) +
  labs(
    x = expression(u[0] ~ "(intercept adjusment)"),
    y = "density"
  )
```



```{r}
ve_speaker |> 
  filter(coef == "voicingvoiced") |> 
  ggplot(aes(r_speaker, reorder(speaker, r_speaker, median))) +
  geom_vline(xintercept = 0) +
  stat_halfeye()
```

```{r}
ve_speaker |> 
  filter(coef == "voicingvoiced") |> 
  ggplot(aes(exp(bs_voicing), reorder(speaker, bs_voicing, median))) +
  geom_vline(xintercept = 1) +
  stat_halfeye()
```
