---
title: "Trading cues to voicing in NC clusters: vowel nasalisation duration and amplitude"
author: "Stefano Coretta"
bibliography: references.bib
---

## The data

First we attach the necessary packages.

```{r}
#| label: setup

library(tidyverse)
library(brms)
library(tidybayes)
theme_set(theme_light())
```

The data comes from @carignan2021a and @cunha2024. I refer the reader to the papers for more information on the data. The data is split in two files, one with the German data and one with the American and British English data. The data contains measures from audio recordings and fMRI recordings.

```{r}
#| label: data

# German data from Carignan et al 2021
nasal_de <- readRDS("data/nasal_tbl_de.rds") |> 
  mutate(language = "de")

# English data from Cunha et al 2024
nasal_en <- readRDS("data/nasal_tbl_en.rds") |> 
  rename(language = english)
```

The following code cleans the data and joins the two tibbles into a single tibble `nasal`. The comments explain the meaning of each new column and between parenthesis it says whether the calculations use acoustic, articulatory or both types of measurements.

```{r}
#| label: nasal

nasal <- bind_rows(nasal_en, nasal_de) |>
  
  # Drop NA values in the specified columns
  drop_na(velumopening_maxvel_on, velumopening_maxvel_off) |>
  
  mutate(
    # Amplitude of nasalisation within the vowel (articulatory)
    nas_ampl = velum2US_Vokal_max,
    
    # Pseudo-gestural nasalisation onset and offset (articulatory)
    art_nas_on = velumopening_nucleus_on - ((velumopening_nucleus_on - velumopening_maxvel_on) / 2),
    art_nas_off = velumopening_nucleus_off + ((velumopening_maxvel_off - velumopening_nucleus_off) / 2),

    # Pseudo-gestural nasalisation duration within the vowel and within the coda
    # (acoustic/articulatory)
    voweln_dur = (Vokal_off - art_nas_on) * 1000,
    codan_dur = (art_nas_off - Vokal_off) * 1000,

    # Proportion of nasalisation in the vowel and coda (acoustic/articulatory)
    voweln_prop = voweln_dur / (Vokal_dur * 1000),
    codan_prop = codan_dur / (Coda_dur * 1000),

    # Factorisation
    vow_lang = factor(interaction(vowel, language, sep = "_")),
    subj_lang = interaction(subject, language, sep = "_"),
    language = factor(language, levels = c("de", "br", "us")),
    language = recode_factor(language, "de" = "German", "br" = "British English", "us" = "American English")
  ) |> 
  
  # Select necessary columns
  select(nas_ampl:subj_lang, language, Vokal_off, Coda_off, Coda_dur, nc_voicing, Satz_dur)

```

Here is a preview of the data.

```{r}
#| label: tbl-nasal
#| tbl-cap: "Preview of the data"
#| echo: false

knitr::kable(head(nasal), digits = 2)
```

## Plotting

@fig-vowelndur shows the density plot of the duration of nasalisation within the vowel in the three languages. Due to the nature of the measurements, some durations are negative (albeit being impossible in principle, nasalisation must start when the nasal consonant starts by definition). The right panel shows the same data but filtered to include only positive duration.

```{r}
#| label: fig-vowelndur
#| fig-cap: "Density plot of the vowel nasalisation duration."
#| fig-subcap: 
#| - "All data."
#| - "Filter data (positive duration)."
#| layout-ncol: 2

nasal |> 
  ggplot(aes(voweln_dur, fill = language)) +
  geom_density() +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Vowel nasalisation duration (ms)"
  ) +
  theme(legend.position = "none")

nasal |> 
  filter(voweln_dur > 0) |> 
  ggplot(aes(voweln_dur, fill = language)) +
  geom_density() +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Vowel nasalisation duration (ms)"
  ) +
  theme(legend.position = "none")
```

@fig-dur-ampl shows a scatter plot of the amplitude of vowel nasalisation vs its duration. The plot suggests a positive correlation between amplitude and duration (Spearman's $\rho =$ `r round(cor(nasal$nas_ampl, nasal$voweln_dur, method = "spearman"), 2)`).

```{r}
#| label: fig-dur-ampl
#| fig-cap: "Scatter plot of vowel nasalisation amplitude vs duration."

nasal |> 
  filter(voweln_dur > 0) |> 
  ggplot(aes(qlogis(nas_ampl), log(voweln_dur))) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Vowel nasalisation amplitude (logistic transf.)",
    y = "Vowel nasalisation duration (logged)"
  )
```


```{r}
#| label: fig-dur-ampl-speaker
#| fig-cap: "Scatter plot of vowel nasalisation amplitude vs duration by speaker."

nasal |> 
  filter(voweln_dur > 0) |> 
  ggplot(aes(qlogis(nas_ampl), log(voweln_prop))) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(subj_lang), scales = "free")
```

We create a filtered tibble with positive vowel duration.

```{r}
#| label: nasal-f

nasal_f <- nasal |> 
  filter(voweln_dur > 0) |>
  mutate(
    # For convenience
    voweln_dur_log = log(voweln_dur),
    nas_ampl_q = qlogis(nas_ampl)
  )
```

@fig-dur-ampl-f is like @fig-dur-ampl but with filtered data (i.e. only positive vowel nasalisation duration). Spearman's $\rho =$ `r round(cor(nasal_f$nas_ampl, nasal_f$voweln_dur, method = "spearman"), 2)`.

```{r}
#| label: fig-dur-ampl-f
#| fig-cap: "Scatter plot of vowel nasalisation amplitude vs duration (filtered data)."

nasal_f |>
  ggplot(aes(nas_ampl_q, voweln_dur_log)) +
  geom_point(alpha = 0.5)
```

## Modelling

```{r}
mod_1 <- brm(
  nc_voicing ~ voweln_prop + nas_ampl_q + Satz_dur +
    (voweln_prop + nas_ampl_q | language/subj_lang),
  data = nasal_f,
  family = bernoulli(),
  seed = 6183,
  cores = 4,
  file = "data/cache/mod_1f"
)
```

```{r}
summary(mod_1, prob = 0.9)
```

```{r}
mod_1_draws <- as_draws_df(mod_1)
```

```{r}
mod_1_r <- mod_1 |> 
  spread_draws(`r_language:subj_lang`[condition,term]) |> 
  pivot_wider(names_from = term, values_from = `r_language:subj_lang`) |> 
  mutate(
    language = case_when(
      str_ends(condition, "de") ~ "DE",
      str_ends(condition, "us") ~ "US",
      str_ends(condition, "br") ~ "BR"
    )
  )
```

```{r}
mod_1_r |> 
  ggplot(aes(nas_ampl_q, voweln_prop, colour = language)) +
  geom_point(alpha = 0.5) +
  facet_grid(cols = vars(language))
```

```{r}
mod_1_r |> 
  ggplot(aes(nas_ampl_q, voweln_prop, colour = language)) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(condition))
```

```{r}
mod_1_r |> 
  group_by(language, .draw) |> 
  summarise(
    cor = cor(nas_ampl_q, voweln_prop),
    .groups = "drop_last"
  ) |> 
  summarise_draws(cor, ~quantile(.x, c(0.025,0.975)) |> round(2))
```

```{r}
mod_1_r |> 
  group_by(condition, language) |> 
  summarise(
    mean_amp = mean(nas_ampl_q), 
    mean_dur = mean(voweln_prop)
  ) |> 
  pivot_longer(mean_amp:mean_dur) |> 
  ggplot(aes(name, value, group = condition)) +
  geom_point() +
  geom_line() +
  facet_grid(cols = vars(language))
```
