---
title: "Trading characteristics of voicing in NC clusters: velum opening onset and amplitude"
author: "Stefano Coretta"
bibliography: references.bib
---

## Overview

This document illustrates how to investigate trading of velum opening timing and its amplitude in individual speakers. The data comes from @carignan2021a and @cunha2024. I refer the reader to the papers for more information on the data. In brief, speakers of German, British or American English had to read words containing VNC sequences while being recorded with functional MRI. The consonant C in the cluster is either voiced or voiceless (for example, *send* vs *sent*).

A nasal consonant requires the velum to be open, but voiceless stops require a tight velum seal to maintain voiceless. It has been proposed that in VNT (where T is a voiceless stop), the velum opening gesture is produced earlier during the vowel than in VND. This should also correspond to greater velum opening at vowel offset in VNT than in VND. Moreover, there should be a negative correlation between opening onset measured from the vowel offset and velum opening magnitude.

A related question is whether speakers, in their productions, trade the time of the velum opening gesture and its amplitude at vowel offset (so that for example, speakers that have an even earlier timing than average also have a greater opening amplitude than average), or whether timing and amplitude are manipulated separately.


## The data

First we attach the necessary packages.

```{r}
#| label: setup

library(tidyverse)
library(brms)
library(tidybayes)
theme_set(theme_light())
```

 The data is split in two files, one with the German data and one with the American and British English data. The data contains measures from audio recordings and fMRI recordings.

```{r}
#| label: data

# German data from Carignan et al 2021
nasal_de <- readRDS("data/nasal_tbl_de.rds") |> 
  mutate(language = "de")

# English data from Cunha et al 2024
nasal_en <- readRDS("data/nasal_tbl_en.rds") |> 
  rename(language = english)
```

The following code cleans the data and joins the two tibbles into a single tibble `nasal`. The comments explain the meaning of each new column and between parenthesis it says whether the calculations use acoustic, articulatory or both types of measurements.

```{r}
#| label: nasal

nasal <- bind_rows(nasal_en, nasal_de) |>
  
  # Drop NA values in the specified columns
  drop_na(velumopening_maxvel_on) |>
  
  mutate(
    # Amplitude of velum opening within the vowel (articulatory)
    vel_max = velum2US_Vokal_max,
    vel_max_q = qlogis(vel_max),
    
    # Onset of velum opening gesture (articulatory)
    vel_on = velumopening_nucleus_on - ((velumopening_nucleus_on - velumopening_maxvel_on) / 2),

    # Onset of velum opening relative to the vowel offset (acoustic/articulatory)
    vel_on_rel = (vel_on - Vokal_off) * 1000,

    # Factorisation
    vow_lang = factor(interaction(vowel, language, sep = "_")),
    subj_lang = interaction(subject, language, sep = "_"),
    language = factor(language, levels = c("de", "br", "us")),
    language = recode_factor(language, "de" = "German", "br" = "British English", "us" = "American English")
  ) |> 
  
  # Select necessary columns
  select(vel_max:subj_lang, language, Vokal_off, Coda_off, Coda_dur, nc_voicing, Satz_dur)

```

Here is a preview of the data.

```{r}
#| label: tbl-nasal
#| tbl-cap: "Preview of the data"
#| echo: false

knitr::kable(head(nasal), digits = 2)
```

## Plotting

@fig-velonrel shows the density plot of the onset of velum opening relative to the vowel offset in the three languages. Due to the nature of the articulatory measurements, some values are positive. We keep the data anyway since this is to illustrate the use of multi-level models.

```{r}
#| label: fig-velonrel
#| fig-cap: "Density plot of the vowel nasalisation duration."

nasal |> 
  ggplot(aes(vel_on_rel, fill = language)) +
  geom_density() +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Vowel nasalisation duration (ms)"
  ) +
  theme(legend.position = "none")
```

@fig-on-ampl shows a scatter plot of the amplitude of velum opening vs its timing. The plot suggests a negative correlation between amplitude and timing (Spearman's $\rho =$ `r round(cor(nasal$vel_on_rel, nasal$vel_max, method = "spearman"), 2)`).

```{r}
#| label: fig-on-ampl
#| fig-cap: "Scatter plot of velum opening timing vs amplitude."

nasal |> 
  ggplot(aes(vel_on_rel, vel_max_q)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Velum opening onset",
    y = "Velum opening amplitude (logistic transf.)",
  )
```


```{r}
#| label: fig-on-ampl-speaker
#| fig-cap: "Scatter plot of velum opening timing vs amplitude by speaker."

nasal |>
  ggplot(aes(vel_on_rel, vel_max_q)) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(subj_lang), scales = "free")
```


## Modelling

We are modelling C voicing as a function of velum opening onset and amplitude. This allows us to see how the relationship with C voicing varies for timing and amplitude in each speaker. The model is a Bernoulli model estimating the probability that the consonant in the VNC sequence is *voiceless*.

```{r}
#| label: mod-1

nasal <- nasal |> 
  mutate(
    vel_on_rel_c = vel_on_rel - mean(vel_on_rel)
  )

mod_1 <- brm(
  nc_voicing ~ vel_on_rel_c + vel_max_q + Satz_dur +
    (vel_on_rel_c + vel_max_q | language/subj_lang),
  data = nasal,
  family = bernoulli(),
  prior = c(
    prior(normal(0, 2.5), class = Intercept),
    prior(normal(0, 1), class = b),
    prior(student_t(3, 0, 2.5), class = sd),
    prior(lkj(2), class = cor)
  ),
  seed = 6183,
  cores = 4,
  file = "data/cache/mod_1"
)
```

The model was not converging with default priors, so I set weakly informative priors. That produces 20 divergent transitions after warmup, which I am fine with. Here is the model summary.

```{r}
#| label: mod-1-summary

summary(mod_1, prob = 0.9)
```

First, as per the regression coefficient `vel_on_rel_c`, later velum opening onset correspond to a change in probability of a voiceless stop by -012 to +0.07 log-odds. In other words, there is a level of uncertainty on whether later velum opening is linked to a greater probability of a voiceless stop.

On the other hand, as per the regression coefficient `vel_max_q`, greater velum opening amplitude corresponds to a decrease in probability of a voiceless stop by -1.58 to -0.45 log-odds. This is counter the hypotheses that VNT sequences should show greater velum opening amplitude (in fact, it is the opposite).

However, we can ask whether speakers still trade the effects of velum opening timing and amplitude. The relevant parameter is the hyper-parameter `cor(vel_on_rel_c,vel_max_q)` in `~language:subj_lang`. This parameter tells us the correlation between the by-speaker adjustments of the `vel_on_rel_c` and `vel_max_q` effects. In this case, the correlation is quite certainly positive: if a speaker shows a stronger effect of timing they also show a stronger effect of amplitude.

Let's extract the posterior draws.

```{r}
#| label: mod-1-draws

mod_1_draws <- as_draws_df(mod_1)
```

In the following, I extract the draws for the multilevel hyper-parameters of the speakers (`subj_lang`). We will need the `vel_max_q` and the `vel_on_rel_c` draws.

```{r}
#| label: mod-1-r

mod_1_r <- mod_1 |> 
  spread_draws(`r_language:subj_lang`[condition,term]) |> 
  pivot_wider(names_from = term, values_from = `r_language:subj_lang`) |> 
  mutate(
    language = case_when(
      str_ends(condition, "de") ~ "DE",
      str_ends(condition, "us") ~ "US",
      str_ends(condition, "br") ~ "BR"
    )
  )
```

```{r}
#| label: fig-draws-scatter
#| fig-cap: "Scatter plot showing a selection of posterior draws of the by-speaker multilevel hyperparameters for the regression coefficients of velum opening onset and amplitude."

mod_1_r |> 
  ungroup() |> 
  slice_sample(n = 300, by = language) |> 
  ggplot(aes(vel_max_q, vel_on_rel_c)) +
  geom_point(alpha = 0.25) +
  geom_smooth(aes(colour = language), method = "lm") +
  facet_grid(cols = vars(language))
```



```{r}
#| label: tbl-cor
#| tbl-cap: "Mean and 95% CrI of the correlation between posterior draws of the by-speaker multilevel hyperparameters for the regression coefficients of velum opening onset and amplitude."

mod_1_r |> 
  group_by(language, .draw) |> 
  summarise(
    cor = cor(vel_max_q, vel_on_rel_c),
    .groups = "drop_last"
  ) |> 
  summarise_draws(mean, ~quantile(.x, c(0.025,0.975))) |> 
  knitr::kable(digits = 2)
```


```{r}
#| label: fig-r-means
#| fig-cap: "Mean of the z-scores of the posterior draws of the multilevel hyper-parameters of the coefficients of velum opening onset and amplitude, by speaker and language."

mod_1_r |> 
  ungroup() |> 
  mutate(
    vel_max_z = scale(vel_max_q),
    vel_on_z = scale(vel_on_rel_c)
  ) |> 
  group_by(condition, language) |> 
  summarise(
    amplitude = mean(vel_max_z), 
    onset = mean(vel_on_z),
    .groups = "drop"
  ) |> 
  pivot_longer(amplitude:onset) |> 
  ggplot(aes(name, value, group = condition)) +
  geom_point() +
  geom_line() +
  facet_grid(cols = vars(language)) +
  labs(
    x = "Hyper-parameter",
    y = "Mean (z-scores)"
  )
```
