---
title: "Trading cues to voicing in NC clusters: proportion and amplitude"
author: "Stefano Coretta"
bibliography: references.bib
---

## The data

First we attach the necessary packages.

```{r}
#| label: setup

library(tidyverse)
library(brms)
library(tidybayes)
theme_set(theme_light())
```

The data comes from @carignan2021a and @cunha2024. I refer the reader to the papers for more information on the data. The data is split in two files, one with the German data and one with the American and British English data. The data contains measures from audio recordings and fMRI recordings.

```{r}
#| label: data

# German data from Carignan et al 2021
nasal_de <- readRDS("data/nasal_tbl_de.rds") |> 
  mutate(language = "de")

# English data from Cunha et al 2024
nasal_en <- readRDS("data/nasal_tbl_en.rds") |> 
  rename(language = english)
```

The following code cleans the data and joins the two tibbles into a single tibble `nasal`. The comments explain the meaning of each new column and between parenthesis it says whether the calculations use acoustic, articulatory or both types of measurements.

```{r}
#| label: nasal

nasal <- bind_rows(nasal_en, nasal_de) |>
  
  # Drop NA values in the specified columns
  drop_na(velumopening_maxvel_on) |>
  
  mutate(
    # Amplitude of velum opening within the vowel (articulatory)
    vel_max = velum2US_Vokal_max,
    vel_max_q = qlogis(vel_max),
    
    # Onset of velum opening gesture (articulatory)
    vel_on = velumopening_nucleus_on - ((velumopening_nucleus_on - velumopening_maxvel_on) / 2),

    # Onset of velum opening relative to the vowel offset (acoustic/articulatory)
    vel_on_rel = (vel_on - Vokal_off) * 1000,

    # Factorisation
    vow_lang = factor(interaction(vowel, language, sep = "_")),
    subj_lang = interaction(subject, language, sep = "_"),
    language = factor(language, levels = c("de", "br", "us")),
    language = recode_factor(language, "de" = "German", "br" = "British English", "us" = "American English")
  ) |> 
  
  # Select necessary columns
  select(vel_max:subj_lang, language, Vokal_off, Coda_off, Coda_dur, nc_voicing, Satz_dur)

```

Here is a preview of the data.

```{r}
#| label: tbl-nasal
#| tbl-cap: "Preview of the data"
#| echo: false

knitr::kable(head(nasal), digits = 2)
```

## Plotting

@fig-velonrel shows the density plot of the onset of velum opening relative to the vowel offset in the three languages. Due to the nature of the articulatory measurements, some values are positive. We keep the data anyway since this is to illustrate the use of multi-level models.

```{r}
#| label: fig-velonrel
#| fig-cap: "Density plot of the vowel nasalisation duration."

nasal |> 
  ggplot(aes(vel_on_rel, fill = language)) +
  geom_density() +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Vowel nasalisation duration (ms)"
  ) +
  theme(legend.position = "none")
```

@fig-on-ampl shows a scatter plot of the amplitude of velum opening vs its timing. The plot suggests a negative correlation between amplitude and timing (Spearman's $\rho =$ `r round(cor(nasal$vel_on_rel, nasal$vel_max, method = "spearman"), 2)`).

```{r}
#| label: fig-on-ampl
#| fig-cap: "Scatter plot of velum opening timing vs amplitude."

nasal |> 
  ggplot(aes(vel_on_rel, vel_max_q)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Velum opening onset",
    y = "Velum opening amplitude (logistic transf.)",
  )
```


```{r}
#| label: fig-on-ampl-speaker
#| fig-cap: "Scatter plot of velum opening timing vs amplitude by speaker."

nasal |>
  ggplot(aes(vel_on_rel, vel_max_q)) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(subj_lang), scales = "free")
```


## Modelling

We are modelling C voicing as a function of velum opening onset and amplitude. This allows us to see how the relationship with C voicing varies for timing and amplitude in each speaker.

```{r}
#| label: mod-1

nasal <- nasal |> 
  mutate(
    vel_on_rel_c = vel_on_rel - mean(vel_on_rel)
  )

mod_1 <- brm(
  nc_voicing ~ vel_on_rel_c + vel_max_q + Satz_dur +
    (vel_on_rel_c + vel_max_q | language/subj_lang),
  data = nasal,
  family = bernoulli(),
  prior = c(
    prior(normal(0, 2.5), class = Intercept),
    prior(normal(0, 1), class = b),
    prior(student_t(3, 0, 2.5), class = sd),
    prior(lkj(2), class = cor)
  ),
  seed = 6183,
  cores = 4,
  file = "data/cache/mod_1"
)
```

```{r}
#| label: mod-1-summary

summary(mod_1, prob = 0.9)
```

```{r}
mod_1_draws <- as_draws_df(mod_1)
```

```{r}
mod_1_r <- mod_1 |> 
  spread_draws(`r_language:subj_lang`[condition,term]) |> 
  pivot_wider(names_from = term, values_from = `r_language:subj_lang`) |> 
  mutate(
    language = case_when(
      str_ends(condition, "de") ~ "DE",
      str_ends(condition, "us") ~ "US",
      str_ends(condition, "br") ~ "BR"
    )
  )
```

```{r}
mod_1_r |> 
  ggplot(aes(vel_max_q, voweln_prop, colour = language)) +
  geom_point(alpha = 0.5) +
  facet_grid(cols = vars(language))
```

```{r}
mod_1_r |> 
  ggplot(aes(vel_max_q, voweln_prop, colour = language)) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(condition))
```

```{r}
mod_1_r |> 
  group_by(language, .draw) |> 
  summarise(
    cor = cor(vel_max_q, voweln_prop),
    .groups = "drop_last"
  ) |> 
  summarise_draws(cor, ~quantile(.x, c(0.025,0.975)) |> round(2))
```

```{r}
mod_1_r |> 
  group_by(condition, language) |> 
  summarise(
    mean_amp = mean(vel_max_q), 
    mean_dur = mean(voweln_prop)
  ) |> 
  pivot_longer(mean_amp:mean_dur) |> 
  ggplot(aes(name, value, group = condition)) +
  geom_point() +
  geom_line() +
  facet_grid(cols = vars(language))
```
