---
title: "Velum opening onset timing and amplitude in VNC sequences with voiceless and voiced consonants"
author: "Stefano Coretta"
bibliography: references.bib
---

## Overview

This document illustrates how to investigate the relationship of velum opening timing and its amplitude with consonant voicing in VNC sequences in the production of individual speakers. The data comes from @carignan2021a and @cunha2024. I refer the reader to the papers for more information on the data. In brief, speakers of German, British or American English had to read words containing VNC sequences while being recorded with functional MRI. The consonant C in the cluster is either voiced or voiceless (for example, *send* vs *sent*).

A nasal consonant requires the velum to be open, but voiceless stops require a tight velum seal to maintain voiceless. It has been proposed that in VNT (where T is a voiceless stop), the velum opening gesture is produced earlier during the vowel than in VND. This should also correspond to greater velum opening at vowel offset in VNT than in VND, assuming a fixed velum opening gesture.

A related question is whether speakers differ systematically in how strongly they use velum-opening timing and amplitude as production cues to the voicing contrast. In particular, we ask whether the speaker-specific effects of the two cues trade off, such that speakers with a stronger association between one cue and consonant voicing show a weaker association with the other, or whether the two cue effects are positively correlated or vary independently across speakers.


## The data

First we attach the necessary packages.

```{r}
#| label: setup

library(tidyverse)
library(brms)
library(tidybayes)
theme_set(theme_light())
```

 The data is split in two files, one with the German data and one with the American and British English data. The data contains measures from audio recordings and fMRI recordings.

```{r}
#| label: data

# German data from Carignan et al 2021
nasal_de <- readRDS("data/nasal_tbl_de.rds") |> 
  mutate(language = "de")

# English data from Cunha et al 2024
nasal_en <- readRDS("data/nasal_tbl_en.rds") |> 
  rename(language = english)
```

The following code cleans the data and joins the two tibbles into a single tibble `nasal`. The comments explain the meaning of each new column and between parenthesis it says whether the calculations use acoustic, articulatory or both types of measurements.

```{r}
#| label: nasal

nasal <- bind_rows(nasal_en, nasal_de) |>
  
  # Drop NA values in the specified columns
  drop_na(velumopening_maxvel_on) |>
  
  mutate(
    # Amplitude of velum opening within the vowel (articulatory)
    vel_max = velum2US_Vokal_max,
    vel_max_q = qlogis(vel_max),
    
    # Onset of velum opening gesture (articulatory)
    vel_on = velumopening_nucleus_on - ((velumopening_nucleus_on - velumopening_maxvel_on) / 2),

    # Onset of velum opening relative to the vowel offset (acoustic/articulatory)
    vel_on_rel = (vel_on - Vokal_off) * 1000,

    # Factorisation
    vow_lang = factor(interaction(vowel, language, sep = "_")),
    subj_lang = interaction(subject, language, sep = "_"),
    language = factor(language, levels = c("de", "br", "us")),
    language = recode_factor(language, "de" = "German", "br" = "British English", "us" = "American English")
  ) |> 
  
  # Select necessary columns
  select(vel_max:subj_lang, language, Vokal_off, Coda_off, Coda_dur, nc_voicing, Satz_dur)

```

Here is a preview of the data.

```{r}
#| label: tbl-nasal
#| tbl-cap: "Preview of the data"
#| echo: false

knitr::kable(head(nasal), digits = 2)
```

## Plotting

@fig-velonrel shows the density plot of the onset of velum opening relative to the vowel offset in the three languages. Due to the nature of the articulatory measurements, some values are positive. We keep the data anyway since this is to illustrate the use of multi-level models.

```{r}
#| label: fig-velonrel
#| fig-cap: "Density plot of the timing of the velum opening relative to the vowel offset by C voicing."

nasal |> 
  ggplot(aes(vel_on_rel, fill = nc_voicing)) +
  geom_density(alpha = 0.5) +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Velum opening onset (ms from V offset)"
  ) +
  theme(legend.position = "none")
```

```{r}
#| label: fig-velampl
#| fig-cap: "Density plot of the amplitude of the velum opening."

nasal |> 
  ggplot(aes(vel_max_q, fill = nc_voicing)) +
  geom_density(alpha = 0.5) +
  geom_rug(alpha = 0.1) +
  facet_grid(vars(language)) +
  labs(
    x = "Velum opening amplitude (logist. transf.)"
  ) +
  theme(legend.position = "none")
```


## Modelling

We are modelling C voicing as a function of velum opening onset and amplitude. This allows us to see how the relationship with C voicing varies for timing and amplitude in each speaker. The model is a Bernoulli model estimating the probability that the consonant in the VNC sequence is *voiceless*.

```{r}
#| label: mod-1

nasal <- nasal |> 
  mutate(
    vel_on_rel_c = vel_on_rel - mean(vel_on_rel)
  )

mod_1 <- brm(
  nc_voicing ~ vel_on_rel_c + vel_max_q + Satz_dur +
    (vel_on_rel_c + vel_max_q | language/subj_lang),
  data = nasal,
  family = bernoulli(),
  prior = c(
    prior(normal(0, 2.5), class = Intercept),
    prior(normal(0, 1), class = b),
    prior(student_t(3, 0, 2.5), class = sd),
    prior(lkj(2), class = cor)
  ),
  seed = 6183,
  cores = 4,
  file = "data/cache/mod_1"
)
```

The model was not converging with default priors, so I set weakly informative priors. That produces 20 divergent transitions after warmup, which I am fine with. Here is the model summary.

```{r}
#| label: mod-1-summary

summary(mod_1, prob = 0.9)
```

First, as per the regression coefficient `vel_on_rel_c`, later velum opening onset correspond to a change in log-odds of a voiceless stop by -0.12 to +0.07. In other words, there is a level of uncertainty on whether later velum opening is linked to a greater probability of a voiceless stop.

On the other hand, as per the regression coefficient `vel_max_q`, greater velum opening amplitude corresponds to a decrease in log-odds of a voiceless stop by -1.58 to -0.45. This is counter the hypothesis that VNT sequences should show greater velum opening amplitude.

```{r}
#| label: mod-2-cond

conditional_effects(mod_1, "vel_max_q")
conditional_effects(mod_1, "vel_on_rel_c")
```


However, we can ask how the effects of velum opening timing and amplitude on the probability of a voiceless consonant in VNC sequences differ relative to each other in each individual speaker. The relevant parameter is the hyper-parameter `cor(vel_on_rel_c,vel_max_q)` in `~language:subj_lang`. This parameter tells us the correlation between the by-speaker adjustments of the `vel_on_rel_c` and `vel_max_q` effects. In this case, the correlation is quite certainly positive: if a speaker shows a stronger effect of timing they also show a stronger effect of amplitude.

Since the effect of opening amplitude is clearly negative, this means that in speakers with a greater effect of timing, the effect of amplitude is less negative.


Let's extract the posterior draws.

```{r}
#| label: mod-1-draws

mod_1_draws <- as_draws_df(mod_1)
```

In the following, I extract the draws for the multilevel hyper-parameters of the speakers (`subj_lang`). We will need the `vel_max_q` and the `vel_on_rel_c` draws.

```{r}
#| label: mod-1-r

mod_1_r <- mod_1 |> 
  spread_draws(`r_language:subj_lang`[condition,term], b_vel_max_q, b_vel_on_rel_c) |> 
  pivot_wider(names_from = term, values_from = `r_language:subj_lang`) |> 
  mutate(
    language = case_when(
      str_ends(condition, "de") ~ "DE",
      str_ends(condition, "us") ~ "US",
      str_ends(condition, "br") ~ "BR"
    ),
    bs_vel_max = b_vel_max_q + vel_max_q,
    bs_vel_on = b_vel_on_rel_c + vel_on_rel_c
  )
```

```{r}
#| label: fig-draws-scatter
#| fig-cap: "Scatter plot showing a selection of posterior draws of the by-speaker multilevel hyperparameters for the regression coefficients of velum opening onset and amplitude."

mod_1_r |> 
  ungroup() |> 
  slice_sample(n = 300, by = language) |> 
  ggplot(aes(vel_max_q, vel_on_rel_c)) +
  geom_point(alpha = 0.25) +
  geom_smooth(aes(colour = language), method = "lm") +
  facet_grid(cols = vars(language))
```


```{r}
#| label: tbl-cor
#| tbl-cap: "Mean and 95% CrI of the correlation between posterior draws of the by-speaker multilevel hyperparameters for the regression coefficients of velum opening onset and amplitude."

mod_1_r |> 
  group_by(language, .draw) |> 
  summarise(
    cor = cor(vel_max_q, vel_on_rel_c),
    .groups = "drop_last"
  ) |> 
  summarise_draws(mean, ~quantile(.x, c(0.025,0.975))) |> 
  knitr::kable(digits = 2)
```


The speaker-specific effects of velum-opening timing and amplitude are strongly positively correlated (ρ = 0.83, 90% CrI [0.67, 0.94]). Thus, speakers do not show evidence of trading the two articulatory parameters against one another. Instead, the two effects are positively coordinated: speakers who show a relatively stronger association between velum-opening onset and consonant voicing also tend to show a relatively stronger association between velum-opening amplitude and consonant voicing. This suggests systematic individual differences in parameter weighting, with some speakers showing stronger parameter–voicing associations across both dimensions and others showing weaker associations across both.

In sum, individual differences appear to lie partly along a common dimension of parameter strength: speakers with a strong association between earlier velum opening and voicelessness also tend to show a strong association between lower velum-opening amplitude and voicelessness, whereas speakers with a weaker onset association tend also to show a weaker amplitude association. This is illustrated in @fig-bs-scatter.

```{r}
#| label: fig-bs-scatter
#| fig-cap: "Mean speaker-specific effects of velum opening onset and amplitude on the probability of a voiceless stop."

mod_1_r |> 
  ungroup() |> 
  group_by(condition, language) |> 
  summarise(
    bs_vel_on = mean(bs_vel_on),
    bs_vel_max = mean(bs_vel_max),
    .groups = "drop"
  ) |> 
  ggplot(aes(bs_vel_on, bs_vel_max, colour = language, shape = language)) +
  geom_point() +
  labs(
    x = "Onset effect (log-odds)",
    y = "Amplitude effect (log-odds)"
  )
```